NumPy---Array Programming in Python

### 2020-01-29 body.tex
A few more changes

Similar examples can be found in physics, earth science, biology,
and more.
Researchers in these fields depend on NumPy being correct, reliable,
and repeatable.
The rich ecosystem of tools building on NumPy and the larger community
of scientists and programmers using those tools mean that NumPy has a
responsibility to maintain considerable stability.
At times, this conflicts with the need to innovate;
resolving it requires balancing trade-offs.

% FIXME: this should be expanded
Over the last few years there has been a massive increase in the use
of scientific Python due in part to the rapid growth of data science,
machine learning, and artificial intelligence.  
As a result, NumPy and its API has become ubiquitous.

@@

% Ralf: we should describe our role at the foundation of the stack
% i was thinking something along these lines ():
NumPy provides in-memory, multidimensional, homogeneously typed
(single pointer + strided) arrays on CPUs.
It supports many types of CPUs (e.g., POWER and ARM) with similar
performance as native C code.
While NumPy suits most user's needs, it cannot support every specialized need of the community;
due to its in-memory data model, it is limited in its ability to work with very
large datasets, datasets split across multiple systems, and computations that
require graphics processing units (GPUs), tensor processing units (TPUs) or
other specialized hardware for performance reasons.

@@

% Ralf: I am not sure where this goes, but it would be good to have something
% about dtype limitation in this section and i also think there should
% be some discussion of missing values somewhere
% (it is mentioned below, but not discussed)
While it supports a wide range of datatypes, it does not have good support
for missing values. % may be worth saying a few sentences about missing values
% if we want to discuss Pandas, perhaps we could do so here (i.e, missing values)

@@

% should we hint at some of the community interaction part
% they have helped us develop the neps and e.g.,
% https://github.com/pytorch/pytorch/issues/22402
To facilitate \emph{interoperability}, the NumPy provides
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
NumPy, in turn, dispatches operations to the originating library, as required.
% FIXME: expand to 3 or more sentences discussing current efforts and
% issues with those, how did adoption go, what is being done next
%\url{https://numpy.org/neps/nep-0037-array-module.html}
Versions of these protocols have been successfully deployed.
% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
% possible to create distributed GPU arrays, enabling .... [find a reference to
% work by Peter Entschev]".
% Best reference: https://ep2019.europython.eu/talks/fX8dJsD-distributed-multi-gpu-computing-with-dask-cupy-and-rapids/
% Below an attempt to start on this:
Thanks to these developments users of Dask and NumPy arrays are now capable
of scaling their computation to the GPU using Dask's CuPy integration with
only minimal code changes.
% spelled out on this page: https://docs.dask.org/en/latest/gpu.html
% The video cited embedded there could be a citation:i
% https://on-demand-gtc.gputechconf.com/gtcnew/sessionview.php?sessionName=s9198-dask+and+v100s+for+fast%2c+distributed+batch+scoring+of+computer+vision+workloads
However, general adoption has not yet been achieved mainly for compatibility
concerns.
% explain a bit what "compatibility concerns" means ...
Thus, we are actively addressing these concerns by revising and extending
these protocols with the aim of wide interoperability that spans the whole
scientific Python ecosystem up to and including domain libraries.

@@

Subclasses of NumPy often exist due to difficulty in constructing richer data
types, such as quantities with physical units \cite{astropy,Goldbaum2018,pint},
geometrical objects \cite{pygeos}, and missing numbers. % pandas: categoricals, IP addresses?
To improve \emph{extensibility}, we are currently overhauling the data type
system to make it easier to implement data types both in Python and C, and to
support these and other applications.
The widespread adoption of rich, user defined, data types can allow more
concise, feature rich scientific computing.
By making these user implemented data types a fundamental design of NumPy
these tools will enabled and interact naturally with different libraries.
% need to add a bit of clarity on what "enable and interact naturally with different libraries" means.
%
%  1. I expect multiple user dtypes can interact (naturally); 2. libraries such
%  as sklearn (or dask) might "just work" with e.g. physical Units (and
%  Categoricals, which they probably really would like to use e.g. for labels);
%  3. Things being inside a dtype rather than subclass is a better user
%  experience.

@@

Over time the role of the NumPy has evolved. It is no longer just the
foundational array library underlying the scientific Python ecosystem,
but it is also a standard API for tensor computation and a
central coordinating mechanism between array types and technologies.

>>>>

Similar examples can be found in physics, earth science, biology,
and more.
% FIXME: this should be expanded 
Over the last few years there has been a massive increase in the use
of scientific Python due in part to the rapid growth of data science,
machine learning, and artificial intelligence.  
As a result, NumPy and its API has become ubiquitous.

@@

% Ralf: we should describe our role at the foundation of the stack
% i was thinking something along these lines ():
NumPy provides in-memory, multidimensional, homogeneously typed
(i.e., single pointer and strided) arrays on CPUs.
It supports many types of CPUs (e.g., POWER and ARM) with similar
performance as native C code.
% a brief word about accelerators perhaps?
% end with how this fits the bill for a huge number of users

While NumPy suits most user's needs, it cannot support every specialized need
of the community; due to its in-memory data model, it is limited in its ability
to work with very large datasets, datasets split across multiple systems, and
computations that require graphics processing units (GPUs), tensor processing
units (TPUs) or other specialized hardware for performance reasons.

@@

% Ralf: I am not sure where this goes, but it would be good to have something
% about dtype limitation in this section and i also think there should
% be some discussion of missing values somewhere
% (it is mentioned below, but not discussed)
While it supports a wide range of datatypes, it does not have good support
for missing values or categorical data. % may be worth saying a few sentences about missing values
% if we want to discuss Pandas, perhaps we could do so here (i.e, missing values)
% maybe we need to rename the section, but this could anticipate
% the dtypes discussion below

@@

% should we hint at some of the community interaction part
% they have helped us develop the neps and e.g.,
% https://github.com/pytorch/pytorch/issues/22402
To facilitate \emph{interoperability}, NumPy provides
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
NumPy, in turn, dispatches operations to the originating library, as required.
% FIXME: expand to 3 or more sentences discussing current efforts and
% issues with those, how did adoption go, what is being done next
%\url{https://numpy.org/neps/nep-0037-array-module.html}
Versions of these protocols have been successfully deployed.
% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
% possible to create distributed GPU arrays, enabling .... [find a reference to
% work by Peter Entschev]".
% Best reference: https://ep2019.europython.eu/talks/fX8dJsD-distributed-multi-gpu-computing-with-dask-cupy-and-rapids/
% Below an attempt to start on this:
Thanks to these developments users of Dask and NumPy arrays are now capable
of scaling their computation to the GPU using Dask's CuPy integration with
only minimal code changes.
% spelled out on this page: https://docs.dask.org/en/latest/gpu.html
% The video cited embedded there could be a citation:i
% https://on-demand-gtc.gputechconf.com/gtcnew/sessionview.php?sessionName=s9198-dask+and+v100s+for+fast%2c+distributed+batch+scoring+of+computer+vision+workloads
However, general adoption has not yet been achieved mainly for compatibility
concerns.
% explain a bit what "compatibility concerns" means ...
Thus, we are actively addressing these concerns by revising and extending
these protocols with the aim of wide interoperability that spans the whole
scientific Python ecosystem up to and including domain libraries.

@@

Subclasses of NumPy often exist due to difficulty in constructing richer data
types, such as quantities with physical units \cite{astropy,Goldbaum2018,pint},
geometrical objects \cite{pygeos}, and missing values. % pandas: categoricals, IP addresses?
To improve \emph{extensibility}, we are currently overhauling the data type
system to make it easier to implement data types both in Python and C, and to
support these and other applications.
The widespread adoption of rich, user defined, data types can allow more
concise, feature rich scientific computing.
By making these user implemented data types a fundamental design of NumPy
these tools will enabled and interact naturally with different libraries.
% need to add a bit of clarity on what "enable and interact naturally with different libraries" means.
%
%  1. I expect multiple user dtypes can interact (naturally); 2. libraries such
%  as sklearn (or dask) might "just work" with e.g. physical Units (and
%  Categoricals, which they probably really would like to use e.g. for labels);
%  3. Things being inside a dtype rather than subclass is a better user
%  experience.

@@

Over time the role of the NumPy has evolved.
It began as an attempt to add an array object to Python
and became the foundation of a rich ecosystem of tools.
Now a large amount of scientific work depends on NumPy
being correct, fast, and stable.
Yet, it is no longer just the foundational array library underlying the
scientific Python ecosystem; it is also a standard API for tensor
computation and a central coordinating mechanism between array types and
technologies.
